---
title: "Reproducible Research: Peer Assessment 1"
author: "moocic"
output: 
html_document:
keep_md: true
---

<br>

## Loading and preprocessing the data

The variables included in this dataset are: `steps` number of steps taking in a 5-minute
interval (missing values are coded as NA); `date` the date on which the measurement was taken
in YYYY-MM-DD format; `interval` identifier for the 5-minute interval in which measurement
was taken.

```{r data, echo = TRUE}
temp <- tempfile()
download.file("https://github.com/moocic/RepData_PeerAssessment1/raw/master/activity.zip", temp)
dataset <- read.csv(unz(temp, "activity.csv"))
unlink(temp)
dataset$date <- as.Date(dataset$date)
head(dataset); str(dataset)
```

<br>

## What is mean total number of steps taken per day?

```{r steps, echo = TRUE}
comp <- complete.cases(dataset)
totalstepsday <- aggregate(dataset[comp, 1],
                           by = list(dataset$date[comp]),
                           FUN = sum,
                           na.rm = TRUE)
totalstepsday_mean <- paste(round(mean(totalstepsday$x), 2))
totalstepsday_median <- median(totalstepsday$x)
totalstepsday_min <- min(totalstepsday$x)
totalstepsday_max <- max(totalstepsday$x)
par(lwd = 2)
hist(totalstepsday$x,
     xlab = "Total number of steps taken per day",
     main = "Histogram",
     ylim = c(0,40))
text(1000, 30,
     paste("Mean =", totalstepsday_mean,
           "\nMedian =", totalstepsday_median),
     pos = 4,
     col = "darkgrey",
     cex = 0.8)
```

The total steps taken per day vary from `r totalstepsday_min` to `r totalstepsday_max`, with a
mean of `r totalstepsday_mean` and a median of `r totalstepsday_median`.

<br>

## What is the average daily activity pattern?

```{r pattern, echo = TRUE}
pattern <- aggregate(dataset[comp, 1],
                     by = list(dataset$interval[comp]),
                     FUN = sum,
                     na.rm = TRUE)
maximum <- which.max(pattern$x)
ymax <- pattern$x[maximum]
xmax <- pattern$Group.1[maximum]
par(lwd = 1.5)
plot(pattern$Group.1,
     pattern$x,
     type = "l",
     xlab = "5-minute interaval (in minutes)",
     ylab = "Average number of steps, averaged across all days")
points(xmax, ymax,
       pch = 16,
       col = "darkgrey")
text(xmax, ymax,
     paste("Maximum at", xmax, "minutes"),
     pos = 4,
     cex = 0.7,
     col = "darkgrey")
```

The 5-minute interval which, on average across all the days in the dataset, contains the
maximum number of steps, was the `r maximum`th 5-minute interval, at `r xmax` minutes, with an
average of `r ymax` steps.

<br>

## Imputing missing values

```{r missings, echo = TRUE}
steps_missings <- sum(is.na(dataset$steps))
row_missings <- sum(!complete.cases(dataset))
percentage <- round(row_missings/nrow(dataset)*100,2)
data.frame(steps_missings, row_missings, percentage)
```

From the observations in the data set, `r row_missings` (`r paste (percentage, "%")`) had
missing values (all in the `steps` variable).


```{r imputing, echo = TRUE}
dailymean <- aggregate(dataset[comp, 1],
                       by = list(dataset$date[comp]),
                       FUN = mean,
                       na.rm = TRUE)
colnames(dailymean) <- c("date", "dailymean")
minutesmean <- aggregate(dataset[comp, 1],
                         by = list(dataset$interval[comp]),
                         FUN = mean,
                         na.rm = TRUE)
colnames(minutesmean) <- c("interval", "minutesmean")
dataset_nona <- merge(dataset, dailymean, by = "date",
                      all = TRUE)
dataset_nona <- merge(dataset_nona, minutesmean, by = "interval",
                      all = TRUE)
dataset_nona <- dataset_nona[order(dataset_nona[, 2], dataset_nona[, 1]), ]
dataset_nona$stepsbefore <- c(dataset_nona$steps[-nrow(dataset_nona)], NA)
dataset_nona$stepsafter <- c(NA, dataset_nona$steps[-1])
naindex <- is.na(dataset_nona$steps)
dataset_nona[naindex, 3] <- rowMeans(dataset_nona[naindex, 4:7],
                                     na.rm = TRUE)
head(dataset_nona); str(dataset_nona)
```

The missing values in the `steps` variable were replaced by the mean of: average steps per day,
average steps per 5-minute interval, steps taken in previous the 5-minute interval and steps taken
in the later 5-minute interval (if data avaiable).

```{r steps_nona, echo = TRUE}
totalstepsday_nona <- aggregate(dataset_nona[, 3],
                                by = list(dataset_nona$date),
                                FUN = sum,
                                na.rm = TRUE)
totalstepsday_mean_nona <- paste(round(mean(totalstepsday_nona$x), 2))
totalstepsday_median_nona <- paste(round(median(totalstepsday_nona$x), 2))
totalstepsday_min_nona <- min(totalstepsday_nona$x)
totalstepsday_max_nona <- max(totalstepsday_nona$x)
par(lwd = 2)
hist(totalstepsday_nona$x,
     xlab = "Total number of steps taken per day",
     main = "Histogram",
     ylim = c(0,40),
     border = "blue")
par(lwd = 1)
hist(totalstepsday$x,
     xlab = "Total number of steps taken per day",
     main = "Histogram",
     ylim = c(0,40),
     border = "red",
     lty = 2,
     add = TRUE)
text(1000, 30,
     paste("Mean =", totalstepsday_mean_nona,
           "\nMedian =", totalstepsday_median_nona),
     pos = 4,
     col = "blue",
     cex = 0.8)
text(1000, 20,
     paste("Mean =", totalstepsday_mean,
           "\nMedian =", totalstepsday_median),
     pos = 4,
     col = "red",
     cex = 0.8)
legend(16000, 30,
       c("Missing values imputed", "Missing values removed"),
       col = c("blue", "red"),
       lty = c(1, 2),
       lwd = c(2, 1),
       bty = "n",
       cex = 0.8)
```

Although the mean and the median total number of steps taken per day did not change greatly,
the number of observations around the mean and median increased by imputing missing data on the
estimates of the total daily number of steps.

<br>

## Are there differences in activity patterns between weekdays and weekends?

```{r differences, echo = TRUE}
dataset_nona$week <- as.numeric(format(dataset_nona$date, "%w"))
dataset_nona$weekdayend <- as.factor(ifelse(dataset_nona$week==0 | dataset_nona$week==6,
                                            "weekend",
                                            ifelse(0 < dataset_nona$week & dataset_nona$week < 6,
                                                   "weekday",
                                                   NA)))
pattern_weekend <- aggregate(dataset_nona[dataset_nona$weekdayend=="weekend", 3],
                             by = list(dataset_nona$interval[dataset_nona$weekdayend=="weekend"]),
                             FUN = mean,
                             na.rm = TRUE)
pattern_weekday <- aggregate(dataset_nona[dataset_nona$weekdayend=="weekday", 3],
                             by = list(dataset_nona$interval[dataset_nona$weekdayend=="weekday"]),
                             FUN = mean,
                             na.rm = TRUE)
maximum_week <- max(c(pattern_weekend$x, pattern_weekday$x))
par(mfrow=c(3,1),
    mai=c(0.3,2,0.3,0.3))
plot(pattern_weekend$Group.1,
     pattern_weekend$x,
     type = "l",
     xlab = "5-minute interaval (in minutes)",
     ylab = "Average number of steps,\non weekend days",
     ylim = c(0, maximum_week*1.1),
     col = "green")
text(2000, 200,
     "Weekend days",
     col = "green",
     cex = 0.8)
plot(pattern_weekday$Group.1,
     pattern_weekday$x,
     type = "l",
     xlab = "5-minute interaval (in minutes)",
     ylab = "Average number of steps,\non weekday days",
     ylim = c(0, maximum_week*1.1),
     col = "purple")
text(2000, 200,
     "Weekday days",
     col = "purple",
     cex = 0.8)
plot(pattern_weekend$Group.1,
     pattern_weekend$x,
     type = "l",
     xlab = "5-minute interaval (in minutes)",
     ylab = "Average number of steps",
     ylim = c(0, maximum_week*1.1),
     col = "green")
lines(pattern_weekday$Group.1,
      pattern_weekday$x,
      col = "purple")
text(2000, 200,
     "Weekend days",
     col = "green",
     cex = 0.8)
text(2000, 150,
     "Weekday days",
     col = "purple",
     cex = 0.8)
```
<br>
<br>
Globally the patterns are similar. However, on average there tend to be more steps taken in mornings on weekdays than on weekends, and more steps taken in the afternoon on weekends than
on weekdays.